Genome Function Phylogenetics

### 2019-04-25 final_project/paper/Paper.Rmd
Added Parsimony method and result to paper

## The Gene Ontology
Historically, the function/role that a gene plays in an organism has always been described in natural language, however the researcher characterizing that gene deemed best.
While this is nice to read, it is not very useful if you want to do computation on it, as computers are (still) horrible at understanding natural language and determining the structure in meaning behind the words.
Additionally, different people will describe the same thing with different words, which has the potential for misunderstanding.

@@

## Data
Annotating genes with their functions can be done experimentally (e.g. by knocking out a certain gene and seeing what processes in the cell are affected), but that is a time-consuming and expensive process, so methods have been developed that try and predict the function of a given gene.
Our lab has developed such a pipeline called GOMAP which combines different prediction approaches and is able to generate high-confidence and very extensive Functional Annotations in a reproducible manner [@Wimalanathan2018].
We have been applying this pipeline to whole-genome assemblies of different plant species and generated functional annotations for every gene in each genome.
These annotation sets are (or will be shortly) available from  
\scriptsize https://dill-picl.org/projects/gomap/gomap-datasets \normalsize.

## Method
Starting point of our method are the functional annotation sets, one for each genome, which annotate every gene in the genome with one or more GO terms.
In more mathematical terms the genome annotation set is a list of tuples $(G, T)$ with $G \in$ Genes in that genome and $T \in$ Terms in the Gene Ontology.
\marginnote[-5em]{
\texttt{Gene\ \ \ \ \ \ \ \ \ GO Term\\
Os01g0601625	GO:0050896\\
Os01g0601625	GO:0016021\\
Os01g0601625	GO:0016301\\
Os01g0601651	GO:0003677\\
Os01g0601651	GO:0009699\\
Os01g0601651	GO:0050790\\
Os01g0601651	GO:0050794\\
Os01g0601651	GO:0050896\\
Os01g0601675	GO:0007275\\
Os01g0601675	GO:0016310\\
Os01g0601675	GO:0050789\\
...
}
}

@@

## Result

The phylogram is displayed in Figure \ref{fig:result}.
4 of the 5 maize assemblies are grouped together in a clade, while the last one is an outlier to the remaining grasses and sits on a very long branch.
In the remaining plants, cotton is an outlier to the legumes, which are all grouped in a single clade.

@@

## Discussion

>>>>

# The Gene Ontology
Historically, the function/role that a gene plays in an organism has always been described in natural language, however the researcher characterizing that gene deemed best.
While this is nice to read, it is not very useful if you want to do computation on it, as computers are (still) horrible at understanding natural language and determining the structure in meaning behind the words.
Additionally, different people will describe the same thing with different words, which has the potential for misunderstanding.

@@

# Data
Annotating genes with their functions can be done experimentally (e.g. by knocking out a certain gene and seeing what processes in the cell are affected), but that is a time-consuming and expensive process, so methods have been developed that try and predict the function of a given gene.
Our lab has developed such a pipeline called GOMAP which combines different prediction approaches and is able to generate high-confidence and very extensive Functional Annotations in a reproducible manner [@Wimalanathan2018].
We have been applying this pipeline to whole-genome assemblies of different plant species and generated functional annotations for every gene in each genome.
These annotation sets are (or will be shortly) available from  
\scriptsize https://dill-picl.org/projects/gomap/gomap-datasets \normalsize.

# Method
I am using two different tree building approaches, one is a distanced based method and the other a parsimony one.

## Distance Based
Starting point of our analysis are the functional annotation sets, one for each genome, which annotate every gene in the genome with one or more GO terms.
In more mathematical terms the genome annotation set is a list of tuples $(G, T)$ with $G \in$ Genes in that genome and $T \in$ Terms in the Gene Ontology.
\marginnote[-5em]{
\texttt{Gene\ \ \ \ \ \ \ \ \ GO Term\\
Os01g0601625	GO:0050896\\
Os01g0601625	GO:0016021\\
Os01g0601625	GO:0016301\\
Os01g0601651	GO:0003677\\
Os01g0601651	GO:0009699\\
Os01g0601651	GO:0050790\\
Os01g0601651	GO:0050794\\
Os01g0601651	GO:0050896\\
Os01g0601675	GO:0007275\\
Os01g0601675	GO:0016310\\
Os01g0601675	GO:0050789\\
...
}
}

@@

## Parsimony Based

Like in the previous method, we again start by enriching all taxa sets $(G,T)$ with their ancestor terms and discarding the gene association: $S = \bigcup_{i=1}^{x}(T_i \cup A_{i})$.

Instead of using these sets for distance calculation, we combine them into a big binary matrix that displays which terms are present in which set:

\begin{table}[]
\begin{tabular}{lllllll}
Taxon       & \scriptsize{GO:0016021} & \scriptsize{GO:0009699} & \scriptsize{GO:0050794} & \scriptsize{GO:0050789} & \scriptsize{GO:0060739} & ... \\
\textit{G. max}      & 1          & 0          & 1          & 0          & 1          &     \\
\textit{T. aestivum} & 1          & 1          & 0          & 1          & 1          &     \\
...         &            &            &            &            &            &    
\end{tabular}
\end{table}

This matrix was then used as the input for `pars` from the PHYLIP package to find the maximally parsimonious tree.

# Results

The phylogram created by the distance based method is displayed in Figure \ref{fig:result}, the maximum parsimony tree in figure \ref{fig:parsimony}.

@@

\begin{figure*}[htbp]
\includegraphics{resources/parsimonyTree.png}
\caption{Phylogram built by looking for the maximally parsimonious tree (total of 7780 changes). Manually rooted between monocots and dicots.}
\label{fig:parsimony}
\end{figure*}

Both trees were combined into a tanglegram (see figure \ref{fig:tanglegram}).

\begin{figure*}[htbp]
\includegraphics{resources/tanglegram.png}
\caption{Tanglegram of the distance based and maximum parsimony tree (both manually rooted).}
\label{fig:tanglegram}
\end{figure*}

# Discussion
